Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-11-27 leadership_across_cultures_contexts.Rmd
Working on MS.
Minor lable edits to analyses files.

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010 and @kaplan_evolutionary_2009; @von_rueden_roots_2014], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and at all levels of social organization. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have also been relatively under-investigated. Current outstanding questions for evolutionary leaderhip scholars include the degree to which the nature of leadership is shaped by population-level variation and the costs and benefits leaders and followers encur across distinct leadership systems.

Evolutionay scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with specific rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them.

@@

The nature of leaderhsip across human societies and social contexts is highly diverse and the emergence of leaders and the functions they serve are closely tied to social and political structure. Among ethnographically described small-scale societies which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect and greater knowledge and social support to achieve or maintain community influence. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of conflict resolution and collective actions. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers. Representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders.

@@

Defensibility of resources necessarily creates increased oppurtunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on disperrsed, less defesnible resources.

@@

Despite immense variation, there do seem to be some commonalities among the qualities of community leaders. Across human societies, when strong community leaders emerge, they tend to be high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations.

Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement.

@@

The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the oppurtunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners.

Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success.

@@

@@

The unique costs associated with leadership positions have received less attention, relative to the benefits of leadership across disciplines.

Group responsibilities necessarily impose opportunity costs on individual priorities and agendas.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions they serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations. Here, we provide the first broad and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample. We also provide the first comparative empirical view describing the costs and benefits of leadership for both leaders nad followers in an evolutionary biological framework. We also aim to understand how the qualities and functions of leaders vary across social contexts and levels of cultural complexity. We hope these exploratory results will facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and social organization.

@@

The operationalized qualities and functions (See Table X) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Tables X. ZG then followed the same process of coding the entire sample of 1212 text records for the X operationalized variables. DISCUSS IRR. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a dataframe named leader_text2).

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs or benefits in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscelaneous costs or benefits not captured by these categories. See Table X.

@@

To reduce the dimensionality of the `r length(quality_vars)` quality variables and the `r length(function_vars)` function variables, all of which take binary values (0, 1), we used logistic PCA from the Binary Dimensionality Reduction (logisticPCA) package [@logisticPCA]. Prior to analysis we removed all rows that had no evidence for any variable. We optimized rank, $k$, and $m$ (the value to approximate the saturated model) using cross-validation, per recommendation of the package authors. Results from logistic PCA stand to (1) reveal which measures (coded variables) distinguish leaders and (2) produce a quantitiave measure of leader qualities and leader functions for additional exploratory analyses.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Table X), we computed the percentage of text records that provided support, estimating standard errors using a binomial generalized linear mixed effects model with a random intercept for author nested within culture. See Figure \@ref(fig:varsupport).

@@

Leaders are most commonly receiving benefits in the form of greater access to resources and increased social status. Recieving social services and mating benefits are also relatively frequent benefits of leadership cross-culturally. The costs follow the potential benefits and leaders stand to lose their high social status and greater access to resources. Followers on the other hand most commonly benefit from leadership systems through receiving resources and social services. Mating and fitness benefits are relatively rare for followers. Followers most frequently encur costs in the form of resources lost and in providing social services. For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and matting opportunities. See Figure \@ref(fig:varsupportcb).

@@

To explore potential population variation in the *Prestige vs. Dominance* leader qualities component and the *Mediation vs. Organization* leader functions component, we fit a series of hierarchical linear models at the text-level, using nested random effects of document within culture with these componets as outcome measures.

@@

We drew on the Standard Cross-Cultural Sample (SCCS) database [@murdock_standard_1969] for additional population-level measures, which scores 186 cultures on many qualitative or quantitative variables, such as kinship system, dependence on agriculture, and social stratification. Fifty of these cultures are the same as, or very similar to, cultures in the PSF. Specifically, we use the SCCS measures of community size, population density, and cultural complexity. Using these measures as covariates allows us to determine the degree to which, an aggregate measure of cultural complexity, mode of subsistence, or measures of community size and population dispersal most strongly impact the qualities and functions of leaders. We also include our researcher coded measure of group structure (See Table X) as a predictor.

>>>>

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and at all levels of social organization. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have also been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is shaped by population-level variation and the costs and benefits leaders and followers incur across distinct leadership systems.

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with specific rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them.

@@

The nature of leadership across human societies and social contexts is highly diverse and the emergence of leaders and the functions they serve are closely tied to social and political structure. Among ethnographically described small-scale societies which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect and greater knowledge and social support to achieve or maintain community influence. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of conflict resolution and collective actions. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers. Representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders.

@@

Defensibility of resources necessarily creates increased opportunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on dispersed, less defensible resources.

@@

Despite immense variation, there do seem to be some commonalities among the qualities of community leaders. Across human societies, when strong community leaders emerge, they tend to be high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006]. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations [@silverman_how_1978; @van_vugt_cognitive_2007; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990]. Drawing on cross-national data through the GLOBE project including data from over 15,000 middle managers across 60 countries, @den_hartog_culture_1999 conclude that elements of charismatic and transformational leadership -- leadership styles when leaders express an inclusive vision of future progress, lead through inspiration, and develop initiatives in collaboration -- are universally endorsed across nations and are perceived to be associated with effective leadership. In particular, inducing motivation, having foresight, being encouraging and communicative, trustworthiness, and building self-confidence among followers are suggested to be universal attributes of outstanding leadership in managerial contexts. Other attributes of charismatic leadership found to be culturally contingent included ambitiousness, self-effacing, enthusiasm, and risk-taking [@den_hartog_culture_1999].

Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012]. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement [@grabo_charismatic_2016].

Among societies with institutionalized social stratification, such as chiefdoms and agricultural communities, community leadership is often highly influenced by heredity and formal class structures [@earle_how_1997; @stanish_evolution_2004]. Managerial elites, who maintain influence over community decisions are endowed with a high degree of ideological status associated with the role and kin group [@stanish_evolution_2010]. In such contexts community leaders often lack coercive authority, but maintain strong influence through ideological systems and ritual or religious status and prowess inaccessible to non-elites [@stanish_evolution_2010]. The association of formal community leadership and supernatural abilities is likely an important precursor of institutionalized authority and social stratification.

Within state-level societies including non-industrial kingdoms leadership positions are proscribed and institutionalized. Ruling classes, parties, or political bodies monopolize social and political influence and maintain control through a variety of mechanisms including military force, ideological systems, and through the control of information and propaganda.

Across discplines many theoretical models and empirical findings have implicated high intelligence as an important quality positively associated with effective leadership [@johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009]. Recent evidence from within managerial studies suggests a curvilinear effect in that exceptionally intelligence individuals at high end of the distribution may not be the most effective leaders in contemporary business settings [@antonakis_can_2017].

@@

The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners.

@@

Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel metaanalysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

@von_rueden_leadership_2015 outilne four possible mechanisms by which leaders in small-scale societies might offset the potential costs associated with various leadership roles; leaders may, (1) facilitate collective actions that yield net benefits to themselves or their close kin, (2) claim a fee for their services, such a tax or greater share of returns, (3) recieve reciprocal exchange in other currencies from followers who have benefited from their services, or (4) may gain other social or reproductive benefits by signaling their high quality as a mate and social partner.

@@

Evolutionary models of leadership generally assume the benefits of leadership outweigh the costs, for both leaders and followers. However, the unique costs associated with various leadership positions have received less attention, relative to the benefits of leadership, across disciplines.

Group responsibilities necessarily impose opportunity costs on individual priorities and agendas, given leadership responsibilities require investments of time, energy, and resources that could be invested elsewhere.

Mediating conflicts between individuals can be physically and socially costly. Devising and proposing compromises and optimal outcomes can be cognitively challenging and the solutions leaders propose are unlikely to be equally satisfactory for everyone involved. Therefore, leaders stand to become the targets of animosity and social ridicule. In the absence of state-based systems and institutionalized inequalities, all group members are on relative equal footing and equipped to challenge leaders directly or indirectly using a variety of leveling mechanisms including overt physical aggression [@boehm_purposive_2008]. Physically formidable individuals with strong social networks are likely to be able o bear these costs more-so than their counterparts and unsurprisingly larger body size and high social capital are qualities often associated with leaders in small-scale societies [@glowacki_leadership_2015]. With greater population density and wealth inequality, these risks are expected to concomitantly increase in magnitude.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions leaders serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations. Here, we provide the first broad and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample. We also provide the first comparative empirical view describing the costs and benefits of leadership for both leaders and followers in an evolutionary biological framework. We also aim to explore how the qualities and functions of leaders vary across social contexts and levels of cultural complexity.

In sum, we aim to empirically and systematically answer the following questions: 1) what are the most frequent qualities of leaders across cultures?; 2) what functions are leaders most commonly serving across cultures?; 3) what are the most salient costs and benefits of leadership for both leaders and followers?; and 4) how does variation across these measures relate to population-level variation?

We hope these exploratory results will facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and organization.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of 1212 text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables. DISCUSS IRR. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named leader_text2).

Table: (\#tab:qualfuncvars) Operational definitions for coded leader qualities and functions.

```{r qualfuncvars, echo=FALSE, message=F, warning=FALSE}
library(readr)
qual_func_var_tab<-read_csv('func_qual_definitions.csv', na = 'NA')
library(knitr)
kable(qual_func_var_tab)
```

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs or benefits in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars).

Table: (\#tab:costbenefitvars) Operational definitions for coded benefits and costs for leaders and followers.

```{r costbenefitvars, echo=FALSE, message=F, warning=FALSE, }
library(readr)
cost_benefit_vars_tab<-read_csv('costs_benefits.csv', na = 'NA')
library(knitr)
kable(cost_benefit_vars_tab)
```

@@

To reduce the dimensionality of the `r length(quality_vars)` quality variables and the `r length(function_vars)` function variables, all of which take binary values (0, 1), we used logistic PCA from the Binary Dimensionality Reduction (logisticPCA) package [@logisticPCA]. Prior to analysis we removed all rows that had no evidence for any variable. We optimized rank, $k$, and $m$ (the value to approximate the saturated model) using cross-validation, per recommendation of the package authors. Results from logistic PCA stand to (1) reveal which measures (coded variables) distinguish leaders and (2) produce a quantitative measure of leader qualities and leader functions for additional exploratory analyses.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded variables (see Tables \@ref(tab:qualfuncvars) and \@ref(tab:costbenefitvars) ), we computed the percentage of text records that provided support, estimating standard errors using a binomial generalized linear mixed effects model with a random intercept for author nested within culture. See Figure \@ref(fig:varsupport).

@@

Leaders are most commonly receiving benefits in the form of greater access to resources and increased social status. Receiving social services and mating benefits are also relatively frequent benefits of leadership cross-culturally. The costs follow the potential benefits and leaders stand to lose their high social status and greater access to resources. Followers on the other hand most commonly benefit from leadership systems through receiving resources and social services. Mating and fitness benefits are relatively rare for followers. Followers most frequently incur costs in the form of resources lost and in providing social services. For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and matting opportunities. See Figure \@ref(fig:varsupportcb).

@@

To explore potential population variation in the *Prestige vs. Dominance* leader qualities component and the *Mediation vs. Organization* leader functions component, we fit a series of hierarchical linear models at the text-level, using nested random effects of document within culture with these components as outcome measures.

@@

<!-- We drew on the Standard Cross-Cultural Sample (SCCS) database [@murdock_standard_1969] for additional population-level measures, which scores 186 cultures on many qualitative or quantitative variables, such as kinship system, dependence on agriculture, and social stratification. Fifty of these cultures are the same as, or very similar to, cultures in the PSF. Specifically, we use the SCCS measures of community size, population density, and cultural complexity. Using these measures as covariates allows us to determine the degree to which, an aggregate measure of cultural complexity, mode of subsistence, or measures of community size and population dispersal most strongly impact the qualities and functions of leaders. We also include our researcher coded measure of group structure (See Table X) as a predictor.   -->
